Abstract Spatial Specification of Stencil Computations

### 2016-06-22 paper.tex
intro update

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteratively computes the one-dimensional discrete Laplace
transform (an approximation to a derivative): 
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   a = b
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. (Note, we use the parentheses syntax
\texttt{( )} for
array subscripts rather than the bracket syntax \texttt{[ ]}.

%This stencil computation exhibits statically decideable
%spatial and temporal relationships between $\texttt{a}$ and
%$\texttt{b}$.
In this simple example, the dependencies between between \texttt{a}
and \texttt{b} form a simple spatial relationship which is easily
understood. This spatial relationship determines other aspects of the
program and its efficient implementation: how much ``boundary'' is
needed for the array, the most cache-efficient layout in memory,
the partitioning shape for parallel implementations.

@@

In this work, we introduce a simple specification language for the
spatial and temporal properties of stencil comptuations. The
specifications abstract over the fine grained detail of stencil access
patterns.  In practise, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered (depth=1, dim=1) :: a
!=  stencil dependency a, mutual       :: b
\end{SpecVerbatim}
%
The first line explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.  The second line explains
the temporal relationship between \texttt{b} and \texttt{a}: that
previous time step for \texttt{b} is actually provided by \texttt{a},
and vice versa. This is explained as a mutual dependence between
\texttt{a} and \texttt{b}. In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The 
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. These can drawn pictorially as follows where the horizontal
dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:

\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}

@@

\begin{figure}
\begin{ExmVerbatim}[firstnumber=20]
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+
    gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))-
    (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-
    gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j)))
    /(4.0*delx)

@@

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+
          (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{ExmVerbatim}
(a). Excerpt from the Fortran code,
showing highly-detailed stencil computations. \\

@@

!=   stencil forward(depth=1, dim=1)
           * backward(depth=1, dim=2) :: v
\end{SpecVerbatim}
(b). Inferred stencil specification from CamFort
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

>>>>

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteratively computes the one-dimensional discrete Laplace
transform (an approximation to a derivative): 
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   a = b
   end do
end do
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. Parentheses 
\texttt{( )} are used here for 
array subscripts rather than bracket syntax \texttt{[ ]}.

%This stencil computation exhibits statically decideable
%spatial and temporal relationships between $\texttt{a}$ and
%$\texttt{b}$.
In this simple example, the dependency between \texttt{a}
and \texttt{b} on line 3 forms a simple spatial relationship which is easily
understood. This spatial relationship determines other aspects of the
program and its efficient implementation: how much ``boundary'' is
needed for the array, the most cache-efficient layout in memory,
the partitioning shape for parallel implementations.

@@

In this work, we introduce a simple specification language for the
spatial properties of stencil comptuations. The
specifications abstract over the fine grained detail of stencil access
patterns.  In practise, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered(depth=1, dim=1) :: a
\end{SpecVerbatim}
%
This explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.  
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The 
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
pictorially.

@@

\begin{figure}[t]
\begin{ExmVerbatim}[firstnumber=20]
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+
    gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))-
    (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-
    gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j)))
    /(4.0*delx)

@@

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+
          (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{ExmVerbatim}
(a). Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. \\

@@

!=   stencil forward(depth=1, dim=1)
           * backward(depth=1, dim=2) :: v
\end{SpecVerbatim}
(b). Inferred stencil specification from CamFort

\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
(c). Pictoral representation of the two stencil specifications.
The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}
